Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2005-06-23 paper/metamonads.tex
try to save more space - not sure how effective this was; all these changes could be reverted
a few misc changes, including biblioghraphy
replacing mdo for doM

\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}   
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

@@

In this paper we report on progress of solving these problems using GE
as our running example. Specifically, our contributions:
\vspace*{-2pt}
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
\begin{comment}
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
\end{comment}
    \item Implementation of the |mdo|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code.
    \item Use functor type sharing constraints to encode domain-specific
      knowledge.
\end{itemize}

@@

let l1 f = fun x -> mdo \{ t <-- x; f t\}
let l2 f = fun x y -> mdo \{ tx <-- x; ty <-- y; f tx ty\}

@@

\begin{figure}
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let param_code4 plus one =
  let gen x y = mdo \{ ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) \}
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let ifM' test th el = mdo \{
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc\}
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> \protect\newline
let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> 
if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
\end{figure}
% param_code3 plus one;;

That does not seem like much of an improvement. With the help of
camlp4 pre-processor, we introduce the |mdo|-notation \cite{metamonadsURL},
patterned after the |do|-notation of Haskell. The function
|param_code4|, written in the |mdo|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor will convert the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

@@

\begin{code2}
module type OUTPUT = sig
  type contr  type res
  module D : DETERMINANT   module R : RANK   module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list, ('a,'w) code) monad
end
module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) = struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det   module R = Rank   module P = DiscardPivot
  let make_result b = mdo \{ det  <-- D.fin ();  rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. \}
end
\end{code2}

@@

\begin{figure}
\begin{code2}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) = struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = mdo \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        mdo \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = mdo \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( mdo \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code2}
\end{figure}

@@

The paper \cite{GluckJ98} describes early simple experiments in
\emph{automatic} and manual staging, and the multi-level language
based on an annotated subset of Scheme (which is untyped and has no
imperative features). The generated code requires post-processing to
attain efficiency. Obtaining a generating extension from annotated
code is an exciting area of our future research.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. We are grateful to anonymous reviewers for many
helpful suggestions.

@@

# val resIV5 :
  ('a,

@@

                    (Funct4.IDet)(Funct4.Rank).res)
  code =

@@

           let a_424 = t_408.arr
           and nm_425 = (t_408.n * t_408.m)

@@

            fun i1_428 ->
             fun i2_429 ->

@@

#   val resFA1 :
  ('a,

@@

                       (Funct4.NoDet(Funct4.FloatDomain)).res)
  code =

>>>>

\renewcommand\floatpagefraction{.95}
\renewcommand\topfraction{.95}
\renewcommand\bottomfraction{.95}
\renewcommand\textfraction{.05}   
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

@@

In this paper we report on progress of solving these problems using GE
as our running example. Specifically, our contributions:
\vspace*{-2pt}
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
\begin{comment}
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
\end{comment}
    \item Implementation of the |doM|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code.
    \item Use functor type sharing constraints to encode domain-specific
      knowledge.
\end{itemize}

@@

let l1 f = fun x -> doM \{ t <-- x; f t\}
let l2 f = fun x y -> doM \{ tx <-- x; ty <-- y; f tx ty\}

@@

\begin{figure}
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let param_code4 plus one =
  let gen x y = doM \{ ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) \}
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let ifM' test th el = doM \{
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc\}
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> \protect\newline
let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> 
if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
\end{figure}
% param_code3 plus one;;

That does not seem like much of an improvement. With the help of
camlp4 pre-processor, we introduce the |doM|-notation \cite{metamonadsURL},
patterned after the |do|-notation of Haskell. The function
|param_code4|, written in the |doM|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor will convert the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

@@

\begin{code2}
module type OUTPUT = sig
  type contr  type res
  module D : DETERMINANT   module R : RANK   module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list, ('a,'w) code) monad
end
module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) = struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det   module R = Rank   module P = DiscardPivot
  let make_result b = doM \{ det  <-- D.fin ();  rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. \}
end
\end{code2}

@@

\begin{figure}
\begin{code2}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) = struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = doM \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        doM \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = doM \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( doM \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code2}
\end{figure}

@@

The paper \cite{GluckJ97} describes early simple experiments in
\emph{automatic} and manual staging, and the multi-level language
based on an annotated subset of Scheme (which is untyped and has no
imperative features). The generated code requires post-processing to
attain efficiency. Obtaining a generating extension from annotated
code is an exciting area of our future research.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. We are grateful to anonymous reviewers for many
helpful suggestions.

@@

# val resIV5 : ('a,

@@

                    (Funct4.IDet)(Funct4.Rank).res) code =

@@

           let a_424 = t_408.arr and nm_425 = (t_408.n * t_408.m)

@@

            fun i1_428 -> fun i2_429 ->

@@

# val resFA1 : ('a,

@@

                       (Funct4.NoDet(Funct4.FloatDomain)).res) code =
